---
title: "Climate Change and Sedimentation: Exploring the Impact of Temperature on Sedimentation"
subtitle: "EES 4891/5891: Probability and Statistics for Geosciences"
author: "Josephine Sinwell"
date: 04/22/2025
---


::: {.comment}
You can find my comments on your project in colored boxes like this.
:::


## Setup

Set up R environment by loading the necessary libraries

```{r setup}
#| message: false
#| warning: false
#| echo: true
#| results: "hide"

# options(repos = c(CRAN = "https://cloud.r-project.org/"))
#
# install.packages("devtools")
# devtools::install_local("C:/Users/Josie/OneDrive/Desktop/EES/EES Stats/lipdR-main")

# This is a good way to ensure that the package is installed, without
# the time-consumping step of installing the package every time you 
# render your document.
if (! require(lipdR)) {
  remotes::install_github("nickmckay/lipdR")
  library(lipdR)
}

library(knitr)
library(scales)
library(tidyverse)
library(tidymodels)
library(lipdR)

theme_set(theme_bw())
```

## Introduction

For this project, I will be analyzing a dataset I found on NOAA titled “Adelaid
Tarn, New Zealand Air Temperature Reconstruction over the Last 16,000 Years.”
This study obtained pollen and plant microfossils from a small lake in South
Island, New Zealand, and used it to create a 16,000-year vegetation and climate
reconstruction. I am interested in this dataset because I am planning to study
abroad in New Zealand, and I find it fascinating how pollen and plant fossils
can tell us about past climates and environmental changes, which can then
inform us about our current and future environments.

::: comment
Good introduction, explaining what the data set is, where you found it, and why
it's interesting to you. The context of how this data set relates to your
research interests, and what you hope to learn from it, is nice.
:::

### Data set

The dataset used in this study is titled “Adelaide Tarn, New Zealand Air
Temperature Reconstruction over the Last 16,000 Years,” and was obtained from
the NOAA National Centers for Environmental Information. It was originally
compiled by Jara et al. (2015) as part of a study reconstructing climate and
vegetation dynamics in the South Island of New Zealand. This dataset includes
variables such as age (calendar year before present), temperature (surface air
temperature), and depth. The dataset covers periods of climatic change, finding
that warming periods 13,000-10,000 years ago and 7,000 to 6,000 years ago led
to forest expansion, while cooling periods 10,000-7,000 years ago and the last
3,000 years led to reduced forests. It also found that climate patterns matched
Antarctic and Southern Ocean trends from 15,000 years ago until about 6,000
years ago, and the switch is attributed to El Nino events causing regional
cooling and forest decline. There are 87 rows of data available.

Citation: Jara, Ignacio A.; Newnham, Rewi M.; Vanderfoes, Marcus J.; Foster,
Courtney R.; Lowe, David J.; Wilmshurst, Janet M.; Moreno, Patricio I.;
Renwick, James A.; Homes, Aline M. 2015. Pollen-climate reconstruction from
northern South Island, New Zealand (41°S), reveals varying high- and
low-latitude teleconnections over the last 16 000 years. Journal of Quaternary
Science, 30(8), 817-829. doi: 10.1002/jqs.2818

::: {.comment}
Good description of what the data is and where it comes from.
You're very thorough, giving the variables in the data set,
who gathered it, where the methods were published.

You also give very good context for what kinds of climatic events
the data can be used to study.

The one additional thing that could have been useful would be to give the
URL for the NOAA data set, since it's not obvious from reading the
paper by Jara _et al._ to figure out where a reader could download the
data.
:::


I loaded the dataset using the lipdR package, which allows for reading and
interacting with Linked Paleo Data files. The dataset was downloaded from NOAA
and read into R as a structured list.

```{r}
# ds <- readLipd("C:/Users/Josie/OneDrive/Desktop/EES/EES Stats/AdelaideTarn.Jara.2015 (2).lpd")
ds <- readLipd("data/AdelaideTarn.Jara.2015.lpd")
```


::: {.comment .problem}
It's a good idea not to hard-code paths, because if you try to run
the analysis on a different computer, the data may not be located
in the same place on the new computer.

Here, I put the data file in a subdirectory of your project, so it 
can be loaded wherever the project is. 

With very large data files, many megabytes, it can be a problem to 
store them in your GitHub repository, but this file is only 
4.5 kilobytes, so there's no problem adding it to the Git repository.
:::


To prepare the dataset for analysis, I extracted the measurement tables
containing the core variables from the LiPD structure.

```{r}
paleo_data <- ds$paleoData[[1]]$measurementTable[[1]]
chron_data <- ds$chronData[[1]]$measurementTable[[1]]

temperature <- paleo_data$temperature$values
depth <- paleo_data$depth$values
age_depth <- unlist(chron_data$depth$values)
age_values <- unlist(chron_data$age$values)

interp_age <- approx(x = age_depth, y = age_values, xout = depth)$y
```

```{r}

df <- data.frame(
  temperature = temperature,
  depth = depth,
  age = interp_age
)

head(df)

```


::: {.comment}
Good job, wrangling the data into a tidy `data.frame`.
:::


## Descriptive Statistics

To understand the basic characteristcs of the dataset, I calculated summary
statistics and standard deviations for the three key variables: temperature,
depth, and age.

```{r}
summary(df$temperature)
summary(df$depth)
summary(df$age)

sd(df$temperature)
sd(df$depth)
sd(df$age, na.rm = TRUE)
```

This shows that temperature has a fairly symmetric distribution around 0, but
the slightly lower median could mean a leftwards skew. Depth also looks fairly
symmetric. Age has a wide spread, and is missing 20 values, but the mean and
median are equal. It has a large standard deviation but that is understandable
since the data ranges across thousands of years.

Next, I made some histograms and kernel density plots to begin to visualize the
data.

```{r}
ggplot(df, aes(x = temperature)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15, fill = "lightblue", color = "white") +
  geom_density(color = "darkblue", linewidth = 0.5) +
  labs(title = "Temperature Distribution", x = "Temperature (°C)", y = "Density")

ggplot(df, aes(x = depth)) +
  geom_histogram(aes(y = after_stat(density)), bins = 25, fill = "pink", color = "white") +
  geom_density(color = "darkblue", linewidth = 0.5) +
  labs(title = "Depth Distribution", x = "Depth (cm)", y = "Density")

ggplot(df |> filter(!is.na(age)), aes(x = age)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10, fill = "purple", color = "white") +
  geom_density(color = "darkblue", linewidth = 0.5) +
  labs(title = "Age Distribution", x = "Age (years Before Present)", y = "Density")
```

For the temperature histogram, it is right skewed, most values are towards the
lower end with fewer values by the tail. This suggests that colder temperatures
were seen more frequently, with a few higher temperature anomalies.


::: {.comment}
Good. 

Something else that is interesting is that the distribution is
bimodal (it has two peaks). It's like there's one part of the data
set that's distributed around 0.2 &deg;C, and another part that's 
distributed around a colder temperature, around $-0.4$ &deg;C.
So it's like the temperature jumps from time to time between a
warmer state and a colder state, with some random variation around 
each state.
:::


For the depth histogram, the shape is fairly uniform. This suggests that the
sediment core was sampled relatively evenly throughout the depth, which is
ideal for calculating sedimentation rates and making comparisons.


::: {.comment}
Good. This is exactly right.

You don't expect depth to follow a probability distribution for
a random variable, because samples are taken at reagular depths.
:::


For the age histogram, the shape has several peaks and dips. Thus, the age
values are not evenly distributed– there must be intervals of slow sediment
accumulation and high accumulation. These could indicate shifts in depositional
environments.


::: {.comment}
Good interpretation of the difference between age and depth.
:::


Next, I made a Q-Q plot to visualize normality for each variable.

```{r}
ggplot(df, aes(sample = temperature)) +
  geom_qq() + geom_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Temperature")

# Depth
ggplot(df, aes(sample = depth)) +
  geom_qq() + geom_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Depth")

# Age (filter out NAs)
ggplot(df |> filter(!is.na(age)), aes(sample = age)) +
  geom_qq() + geom_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Age")
```

For the temperature Q-Q plot, the data is roughly normally distributed, the
points generally align with the line. However there is some deviation at the
upper and lower ends.

::: {.comment}
I'd say it's not very normal. It deviates a good deal from the line
near 0, so this shows that the median is greater than the mean, and
there's a significant skew that affects most of the data.

The way the Q-Q plot looks curved near 0 on the x-axis 
(here, it looks like it's curved downward, going from above the line
to below the line), is an indication of a skewed distribution. 
Either an upward or a downward curve near the middle indicates skew.
is a
:::


For the depth Q-Q plot, the data here does not show a normal distributed, with
a significant departure from the red line, especially at the tails. The data is
likely skewed or has outliers.

For the age Q-Q plot, it shows a similar pattern as the depth plot. Large
deviations shown, especially at the higher end.


::: {.comment}
You know from your histograms and kernel-density plots that the depth
and age are not normally distributed, so you don't expect the Q-Q
plots to look normal, and you correctly observe that they are clearly
not normal.
:::


## Research Questions

This study aims to explore how climate variation influenced sediment deposition
over time at Adelaide Tarn. Specifically, I want to investigate the
relationship between temperature (used as an indicator for climate change) and
sedimentation rate, which is derived from changes in depth and age.

My main research question is: How were sedimentation rates affected by climate
change during the period recorded in the sediment core?

Sub-questions I will look at include:

-   Does temperature correlate significantly with sedimentation rate?

-   Does sedimentation rate show significant variability with temperature?


::: {.comment}
These are good, interesting research questions.
:::

## Hypotheses and Tests

### Hypotheses

Main Hypothesis: periods of climatic extremes are associated with increased
sedimentation rates, suggesting that significant climate cooling and warming
may have influenced depositional processes such as erosion or runoff.

Null Hypothesis: there is no significant relationship between sedimentation
rate and temperature.


::: {.comment}
Good hypothesses to use for answering your research questions.
:::


### Methods

First I will calculate sedimentation rates using the age-depth relationship. I
will also clean up the data by removing any non-available data.

```{r}
df <- df |> 
  arrange(age) |>
  mutate(sed_rate = c(NA, diff(depth) / diff(age))) |>
  filter(!is.na(sed_rate), !is.na(temperature), !is.na(age))
```


::: {.comment}
Good way to calculate the sedimentation rate
:::


Then I will check for linear or nonlinear relationships between sedimentation
rate and temperature, using scatterplots and regression models.

```{r}
ggplot(df, aes(x = temperature, y = sed_rate)) +
  geom_point() +
  geom_smooth(method = "lm", color = "purple") +
  labs(
    title = "Relationship Between Temperature and Sedimentation Rate",
    x = "Temperature (°C)",
    y = "Sedimentation Rate (cm/year"
  )
```

We see here that the trend line is relatively flat, which suggests that there
may not be a strong relationship between temperature and sedimentation rate.
There is also a large spread of data points, scattered especially in the higher
temperature range. This indicates a large variability in the sedimentation
rate.


::: {.comment}
So something interesting here is that there are four extreme outliers
from the rest of the data. If you inspect the data frame, you will 
finnd that thesse extreme values are the four shallowest samples,
at depths 20, 25, 30, and 35.


```{r }
#| class.source:  "grade-src"
#| class.output:  "grade-out"
#| class.message: "grade-msg"
#| class.warning: "grade-warn"
#| class.error:   "grade-err"
df |> filter(sed_rate > 0.1)

head(df, 10)
```

So some of what you may be seeing here is that over time, sediment
compacts, especially as new layers are deposited and the weight of the
upper layers compressses lower layers.

You might want to leave the first four points out when you analyze the
rest of your data.
:::

Then I will test whether sedimentation rate is normally distributed by creating
a Q-Q plot and then conducting a Shapiro-Wilk test for normality.

```{r}
ggplot(df, aes(sample = sed_rate)) +
  geom_qq() +
  geom_qq_line()

shapiro.test(df$sed_rate)
```

In the Q-Q plot, there is a significant departure of the datapoints from the
line, particularly at the tails. The sedimentation rate does not seem normally
distributed.

The Shapiro-Wilk test confirms that the sedimentation rate is not normally
distributed. The p value is incredibly small, which confirms the null
hypothesis.


::: {.comment}
It can be interesting to repeat the normality tests after removing
the outlier points


```{r }
#| class.source:  "grade-src"
#| class.output:  "grade-out"
#| class.message: "grade-msg"
#| class.warning: "grade-warn"
#| class.error:   "grade-err"
df |> filter(depth > 35) |> ggplot(aes(sample = sed_rate)) + 
  geom_qq_line() + geom_qq()
```

And we can also look at the Shapiro-Wilk test


```{r }
#| class.source:  "grade-src"
#| class.output:  "grade-out"
#| class.message: "grade-msg"
#| class.warning: "grade-warn"
#| class.error:   "grade-err"
df |> filter(depth > 35) |> pull(sed_rate) |> shapiro.test()

```

The data is still not normally distributed, but it's a lot closer
to a normal distribution than when we included the four shallowest
observations.

:::


At this point, I decided to try to fit my model to a gamma distribution.


::: {.comment .problem}
A subtle misunderstanding here is for a regression analysis, what you
want to test is not the probability distribution of the raw data, but
the distribution of the residuals.

You would fit the data to a linear relationship, 
`sed_rate ~ temperature`, and then analyze the distribution of the 
residuals.

A linear regression can work very well if the raw data is not 
normally distributed, as long as the residuals are reasonably close
to normally distributed.
:::


```{r}
library(MASS, quietly = TRUE)
g_params <- fitdistr(df$sed_rate, densfun = "gamma")
print(g_params$estimate)
```

Now, I made a Q-Q plot to check how well the data is described by a gamma
distribution.

```{r}
ggplot(df, aes(sample = sed_rate)) + 
  geom_qq_line(distribution = qgamma, dparams = g_params$estimate) +
  geom_qq(distribution = qgamma, dparams = g_params$estimate) +
  labs(title = "Q-Q Plot for Sedimentation Rate (Gamma Distribution)",
       x = "Theoretical Quantiles", y = "Sample Quantiles")
```

This fits a little bit better, especially in the lower range, so we will go
with it. There is still deviation in the higher sedimentation ranges however.

### Hypothesis Tests

Next I will apply linear regression to test the relationship between
temperature and sedimentation rate. To do this I will fit a generalized linear
model with Gamma Distribution.

I had to do a bit of research on how to do regression analysis with a gamma
distribution, which is where I learned that you need to use a generalized
linear model and an appropriate link function. Further research led me to
select a log link function as the appropriate link function, as it is commonly
used when the dependent variable is positive and continuous, like my
sedimentation rate.

```{r}
recipe <- recipe(sed_rate ~ temperature, data = df)
model <- linear_reg() |> 
  set_engine("glm", family = Gamma(link = "log"))

wflow <- workflow() |> 
  add_recipe(recipe) |> 
  add_model(model)

fit <- fit(wflow, data = df)
fit |> extract_fit_engine() |> summary()
```


::: {.comment .problem}
It would be worth doing a regular `lm()` fit here, or a `glm()` model
with a linear link instead of a log link.

If you do this and filter out depths shallower than 40 cm, 
you find a statistically significant coefficient for
temperature:


```{r }
#| class.source:  "grade-src"
#| class.output:  "grade-out"
#| class.message: "grade-msg"
#| class.warning: "grade-warn"
#| class.error:   "grade-err"

model_2 <- linear_reg() |> set_engine("glm")
wflow_2 <- workflow() |>
  add_recipe(recipe) |> 
  add_model(model_2)

fit_2 <- fit(wflow_2, data = filter(df, depth > 35 ))
fit_2 |> extract_fit_engine() |> summary()
```

You would also see a significant coefficient if you used your Gamma
model with a log link omitting the four outlier points:


```{r }
#| class.source:  "grade-src"
#| class.output:  "grade-out"
#| class.message: "grade-msg"
#| class.warning: "grade-warn"
#| class.error:   "grade-err"
fit_3 <- fit(wflow, data = df |> filter(depth > 35))
fit_3 |> extract_fit_engine() |> summary()
```


This could have been a good place to try out workflow mapping to 
compare several models. You could check linear and log links, 
and compare `lm` and `glm`. Using $k$-fold cross-validation would let
you test model accuracy without having to make assumptions about whether
the residuals are distributed normally, or on any other specific 
distribution, such as gamma.
:::

For the intercept, which represents the expected value of the sedimentation
rate when temperature is 0, the estimate was -3,004488. Since this is a log
link function, the intercept represents the log-transformed expected
sedimentation rate. The intercept is highly statistically significant, as show
by the very small p-value.

For the temperature, the coefficient indicates that for each 1°C increase in
temperature, the sedimentation rate decreases by approximately 0.78. This
p-value of 0.391, however, is too big to be statistically significant, thus
there is no strong evidence to support that temperature signifcantly affects
sedimentation rate in this model.

For the model fit, the null deviance is 18.607 and the residual deviance is
18.341. Since the deviance after adding the temperature predictor is only
slightly smaller, it suggests that the model with temperature does not really
add anything.


::: {.comment}
Good interpretation of what the model summary says.
:::


### Results and Discussion

To recap, my main hypothesis was that periods of climatic extremes are
associated with increased sedimentation rates, suggesting that significant
climate cooling and warming may have influenced depositional processes such as
erosion or runoff. My null hypothesis was that there is no significant
relationship between sedimentation rate and temperature.

To test this, I used a Gamma GLM with a log link function, focusing on
temperature as the main predictor. This resulted in a p-value of 0.391, which
is not statistically significant. This indicates that temperature does not
significantly influence sedimentation rates in the dataset, meaning that my
initial hypothesis was not supported by the data.

This lack of significance indicates that temperature alone may not explain the
variability in sedimentation rates in this dataset. This could be due to other
factors such as precipitation, vegetation, or human activity possibly having a
more substantial influence on sedimentation rates.


::: {.comment}
The biggest thing that contributes to the lack of significance are
the four outlier points in sedimentation rate, corresponding to the 
shallowest four depths.
:::


## Conclusions

I learned that temperature does not always provide a direct means of
understanding sedimentation rates. More importantly, I gained a lot of hands-on
experience with implementing statistical regression on environmental data,
which has significantly contributed to my learning process in this class. Even
though temperature was an expected predictor, it did not show the significant
effect that I had anticipated, showing first-hand how complex environmental
processes can be.

A reader would likely be interested in the fact that temperature did not have a
significant impact on sedimentation rates, which, as I experienced, may
challenge previous assumptions about the impact climate has on sedimentation.
This also highlights the importance of considering multiple variables when
analyzing environmental data.

For future studies, it might be best to consider some of these other variables,
such as precipitation, vegetation, and human activity. Non-linear regression
might also be helpful, but I do not know enough about it. It might also be
interesting to look at extended time periods, or even just a different time
period. This could add more samples to the dataset and potentially offer a more
accurate analysis. An expansion of the timeframe could also lead to more
generalized studies on large-scale climate trends.


::: {.comment}

### Summary 

This is a very good analysis. It is clear that you learned a lot. 

There were a few places where you could have done better:

1. Drop the first four observations, because the deposition rates are
   a lot larger than any of the rest of the data. I suspect this is 
   because of compaction for the deeper (older) layers. The 
   shallowest layerss may be too young to have had time to fully 
   compact, and also the weight of the overburden contributes to
   compaction, and the shallowest layers don't have as much mass
   overhead to press down on it.
2. You misunderstood the use of normality testing for figuring out the
   best kind of regression. You could have used a simpler regression,
   such as `lm()` or `glm()` with a linear link, and you didn't need
   to specify a gamma distribution.
   
This would have allowed you to do a better hypothesis test, and you
would have found a significant relationship between temperature and
deposition rate. 

Apart from these issues, you did a very nice job with this analysis,
and I thought the way you integrated the text describing what you 
were doing, the code to implement your analysis, and the output
(graphs, and text output), to show the reader the results.

**Grade:** $A-$
:::

